Data Pre-Processing/Cleaning Script

### 2023-04-07 code/processing_code/processingfile_v1.qmd
update

## Training Data
```{r}
training_data_fin <- training_data %>%
  select(!c(Ingredient, `Serotype/Genotype`)) %>%
  filter(grepl("Confirmed", Status))
```
# Data Manipulation
## Training Data
### Cleaning the "Species" Column
```{r}
unique(training_data_fin$Species)
# Separate the rows for coninfections
training_data_fin=training_data_fin%>%
  separate_rows(Species,sep=";")
# Create lists for pathogen types
Pathogen_Bacteria <- c("Bacillus", "Brucella", "Campylobacter", "Clostridium", "Staphylococcus", "Escherichia", "Listeria", "Salmonella", "Vibrio", "Cholera", "Enterococcus", "Bacterium", "Shigella", "Yersinia","Streptococcus")
Pathogen_Parasite <- c("Cryptosporidium", "Cyclospora", "Entamoeba", "Giardia", "Parasite", "Toxoplasma", "Trichinella")
Pathogen_Virus <- c("Astrovirus", "Norovirus", "Rotavirus","Hepatitis","Virus")
Pathogen_Poison <- c("Poison", "Toxin", "Metal","Pesticide","Ciguatoxin","Scombroidtoxin")
# simplify "Escherichia coli, Shiga toxin-producing" to "Escherichia coli"
# Create two new columns: Pathogen_type & Simplified_Etiology
training_data_fin=training_data_fin%>%
  mutate(Pathogen_Type = case_when(
    grepl(paste(Pathogen_Virus,
                collapse="|"), Species,ignore.case = TRUE) ~ "Virus",
    grepl(paste(Pathogen_Bacteria, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Bacteria",
    grepl(paste(Pathogen_Parasite, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Parasite",
    grepl(paste(Pathogen_Poison, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Poison"
    )) %>%
  mutate(Simplified_Etiology = word(Species,1))%>%
      mutate(Simplified_Etiology = case_when(
        Simplified_Etiology == "Escherichia" ~ "E. coli",
        Simplified_Etiology == "Heavy" ~ "Heavy Metals",
        TRUE ~ Simplified_Etiology))
```

### Column for Location
```{r}
# Checking Location Names and Separate Multiple Names
unique(training_data_fin$Location)
training_data_fin=training_data_fin %>%
 separate_rows(Location,sep="; ")
```

@@

```{r}
# unique(analysis_data_fin$`IFSAC Category`)
# unique(training_data_fin$Food)
Desserts=c("Cookie","Cream","Canoli","Cake","Pudding")
Beef=c("Beef","Steak")
Poultry=c("Chicken","Turkey")
Pork=c("Pork","Sausage","Hot Dog","Deli","Spam","Strawberries","Blueberries","Pig","Ribs")
Dairy=c("Milk","Cheese","Parsley")
Crusteaceans=c("Lobster","Crab","Shrimp")
Fish=c("Fish","Tuna","Salmon")
Mollusks=c("Clams","Oysters")
Fruits=c("Tomato","Banana","Melon","Grapes","Mango")
`Grains-Beans`=c("Rice","Peas","Beans","wheat")
`Vegetable-Row-Crops`=c("Lettuce","artichokes","broccoli","asparagus","celeries","spinach")
`Root-underground-vegetables`=c("Roots","carrots","Beets","Onion","Tubers","yams","Garlic","Potatos","taro")
Sprouts=c("alfalfa","mung")
Salad=c("Salad")
# Other=c("Ice","Alcohol","Eggnog")
Game=c("Bear")
Herbs=c("basil","cilantro")
Fungi=c("Mushroom","portabellas")
`Nuts-Seeds`=c("peanuts","almonds","Seasames")
Eggs=c("Egg")
`Seeded-Vegetables`=c("Peppers","Squashes","Cucumbers","Okras","Corn")
`other meat`=c("Lamb")

training_data_fin <- training_data_fin %>%
  mutate(`IFSAC Category` = case_when(
    grepl("Salad",Food, ignore.case = TRUE) ~ "Salad",
    grepl(paste(Desserts, 
                collapse = "|"), Food,ignore.case = TRUE) ~ "Dessert",
    grepl(paste(Poultry,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Poultry",
    grepl(paste(Beef,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Beef",
    grepl(paste(Pork,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Pork",
    grepl(paste(Dairy,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Dairy",
    grepl(paste(Crusteaceans,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Crusteaceans",
    grepl(paste(Fish,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Fish",
    grepl(paste(Mollusks,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Mollusks",
    grepl(paste(Fruits,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Fruits",
    grepl(paste(`Grains-Beans`,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Grains-Beans",
    grepl(paste(`Vegetable-Row-Crops`,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Vegetable-Row-Crops",
    grepl(paste(`Root-underground-vegetables`,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Root-underground-vegetables",
    grepl(paste(Sprouts,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Sprouts",

    grepl(paste(Herbs,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Herbs",
    grepl(paste(Fungi,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Fungi",
    grepl(paste(`Nuts-Seeds`,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Nuts-Seeds",
    grepl(paste(`Seeded-Vegetables`,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Seeded-Vegetables",
    grepl("eggs",Food,ignore.case = TRUE)~"Eggs",
    grepl("Lamb",Food)~"Other Meat",
    TRUE ~ if_else(is.na(Food),"","Other")
   ))%>%
  mutate(`IFSAC Category` = na_if(`IFSAC Category`, ""))%>%
  relocate(`IFSAC Category`, .after = Food)
training_data_fin=training_data_fin%>%
  mutate(`IFSAC Category`=case_when(
    grepl(";",Food)~"Multiple",TRUE~`IFSAC Category`))
```

@@

### Column for Type of Pathogen
```{r}

testing_data_fin <- testing_data_fin %>%
  mutate(Pathogen_Type = case_when(
    grepl(paste(Pathogen_Poison, 
                collapse = "|"), Etiology) ~ "Poison",
    grepl(paste(Pathogen_Bacteria, 
                collapse = "|"), Etiology) ~ "Bacteria",
    grepl(paste(Pathogen_Parasite, 
                collapse = "|"), Etiology) ~ "Parasite",
    grepl(paste(Pathogen_Virus, 
                collapse = "|"), Etiology) ~ "Virus"
    )) %>%
  mutate(Simplified_Etiology = word(Etiology,1)) %>%
      mutate(Simplified_Etiology = case_when(
        Simplified_Etiology == "Escherichia" ~ "E. coli",
        Simplified_Etiology == "Heavy" ~ "Heavy Metals",
        TRUE ~ Simplified_Etiology))
```
###Column for Location
```{r}
testing_data_fin <- testing_data_fin %>%
  rename(Location=Setting)%>%
  mutate(Location=case_when(
    grepl("Restaurant",Location)~"Restaurant",TRUE~Location))%>%
    separate_rows(Location,sep="; ")
testing_data_fin
```

# Drop Rows Containing NA
```{r}
training_data_final=training_data_fin%>%
  drop_na()
testing_data_final=testing_data_fin%>%
  drop_na()

>>>>

## Training Data
```{r}
training_data_fin <- training_data %>%
  select(!c(Ingredient, `Serotype/Genotype`)) %>%
  filter(grepl("Confirmed", Status))
```
# Data Manipulation
## Training Data
### Cleaning the "Species" Column
```{r}
unique(training_data_fin$Species)
# Separate the rows for coninfections
training_data_fin <- training_data_fin %>%
  separate_rows(Species, sep= ";")
# Create lists for pathogen types
Pathogen_Bacteria <- c("Bacillus", "Brucella", "Campylobacter", "Clostridium", "Staphylococcus", "Escherichia", "Listeria", "Salmonella", "Vibrio", "Cholera", "Enterococcus", "Bacterium", "Shigella", "Yersinia","Streptococcus")
Pathogen_Parasite <- c("Cryptosporidium", "Cyclospora", "Entamoeba", "Giardia", "Parasite", "Toxoplasma", "Trichinella")
Pathogen_Virus <- c("Astrovirus", "Norovirus", "Rotavirus","Hepatitis","Virus")
Pathogen_Poison <- c("Poison", "Toxin", "Metal","Pesticide","Ciguatoxin","Scombroidtoxin")
# simplify "Escherichia coli, Shiga toxin-producing" to "Escherichia coli"
# Create two new columns: Pathogen_type & Simplified_Etiology
training_data_fin <- training_data_fin %>%
  mutate(Pathogen_Type = case_when(
    grepl(paste(Pathogen_Virus,
                collapse="|"), Species,ignore.case = TRUE) ~ "Virus",
    grepl(paste(Pathogen_Bacteria, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Bacteria",
    grepl(paste(Pathogen_Parasite, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Parasite",
    grepl(paste(Pathogen_Poison, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Poison"
    )) %>%
  mutate(Simplified_Etiology = word(Species,1))%>%
      mutate(Simplified_Etiology = case_when(
        Simplified_Etiology == "Escherichia" ~ "E. coli",
        Simplified_Etiology == "Heavy" ~ "Heavy Metals",
        TRUE ~ Simplified_Etiology))
```

### Column for Location
```{r}
# Checking Location Names and Separate Multiple Names
unique(training_data_fin$Location)
training_data_fin <- training_data_fin %>%
 separate_rows(Location,sep = "; ")

# Simplify Location
training_data_fin <- training_data_fin %>% mutate(Location = case_when(
  grepl("Buffet", Location, ignore.case = TRUE) ~ "Restaurant",
  grepl("Nursing Home", Location, ignore.case = TRUE) ~ "Nursing Home/Assisted Living Facility",
  grepl("Fast Food", Location, ignore.case = TRUE) ~ "Restaurant",
  TRUE ~ Location))
```

@@

```{r}
# unique(testing_data_fin$`IFSAC Category`)
# unique(training_data_fin$Food)

# Dairy
Desserts <- c("Cookie","Cream","Canoli","Cake","Pudding")
Dairy <- c("Milk","Cheese","Parsley")

# Meat & Protein
Beef <- c("Beef","Steak")
Poultry <- c("Chicken","Turkey")
Pork <- c("Pork", 
          "Sausage",
          "Hot Dog", 
          "Deli", 
          "Spam", 
          "Strawberries", 
          "Blueberries", 
          "Pig", 
          "Ribs")
Crusteaceans <- c("Lobster","Crab","Shrimp")
Fish <- c("Fish","Tuna","Salmon")
Mollusks <- c("Clams","Oysters")
Game <- c("Bear")
`other meat` <- c("Lamb")

# Plant-Based Foods
Fruits <- c("Tomato","Banana","Melon","Grapes","Mango")
`Grains-Beans` <- c("Rice","Peas","Beans","wheat")
`Vegetable-Row-Crops` <- c("Lettuce", "artichokes", "broccoli", "asparagus", 
                           "celeries", "spinach")
`Root-underground-vegetables` <- c("Roots", "carrots", "Beets", "Onion", 
                                   "Tubers", "yams", "Garlic", "Potatos", 
                                   "taro")
Sprouts <- c("alfalfa","mung")
Salad <- c("Salad")

# Other=c("Ice","Alcohol","Eggnog")

Herbs <- c("basil","cilantro")
Fungi <- c("Mushroom","portabellas")
`Nuts-Seeds` <- c("peanuts","almonds","Seasames")
Eggs <- c("Egg")
`Seeded-Vegetables` <- c("Peppers","Squashes","Cucumbers","Okras","Corn")

training_data_fin <- training_data_fin %>%
  mutate(`IFSAC Category` = case_when(
    grepl("Salad", Food, ignore.case = TRUE) ~ "Salad",
    grepl(paste(Desserts, 
                collapse = "|"), Food,ignore.case = TRUE) ~ "Dessert",
    grepl(paste(Poultry,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Poultry",
    grepl(paste(Beef,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Beef",
    grepl(paste(Pork,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Pork",
    grepl(paste(Dairy,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Dairy",
    grepl(paste(Crusteaceans,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Crusteaceans",
    grepl(paste(Fish,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Fish",
    grepl(paste(Mollusks,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Mollusks",
    grepl(paste(Fruits,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Fruits",
    grepl(paste(`Grains-Beans`,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Grains-Beans",
    grepl(paste(`Vegetable-Row-Crops`,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Vegetable-Row-Crops",
    grepl(paste(`Root-underground-vegetables`,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Root-underground-vegetables",
    grepl(paste(Sprouts,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Sprouts",

    grepl(paste(Herbs,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Herbs",
    grepl(paste(Fungi,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Fungi",
    grepl(paste(`Nuts-Seeds`,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Nuts-Seeds",
    grepl(paste(`Seeded-Vegetables`,
                collapse = "|"), Food, ignore.case = TRUE) ~ "Seeded-Vegetables",
    grepl("eggs", Food, ignore.case = TRUE)~"Eggs",
    grepl("Lamb", Food) ~ "Other Meat",
    TRUE ~ if_else(is.na(Food),"","Other")
   ))%>%
  mutate(`IFSAC Category` = na_if(`IFSAC Category`, ""))%>%
  relocate(`IFSAC Category`, .after = Food)

training_data_fin <- training_data_fin %>%
  mutate(`IFSAC Category`= case_when(
    grepl(";", Food) ~ "Multiple",
    TRUE ~ `IFSAC Category`))
```

@@

### Column for Type of Pathogen
```{r}
testing_data_fin <- testing_data_fin %>%
  mutate(Pathogen_Type = case_when(
    grepl(paste(Pathogen_Poison, 
                collapse = "|"), Etiology) ~ "Poison",
    grepl(paste(Pathogen_Bacteria, 
                collapse = "|"), Etiology) ~ "Bacteria",
    grepl(paste(Pathogen_Parasite, 
                collapse = "|"), Etiology) ~ "Parasite",
    grepl(paste(Pathogen_Virus, 
                collapse = "|"), Etiology) ~ "Virus"
    )) %>%
  mutate(Simplified_Etiology = word(Etiology, 1)) %>%
      mutate(Simplified_Etiology = case_when(
        Simplified_Etiology == "Escherichia" ~ "E. coli",
        Simplified_Etiology == "Heavy" ~ "Heavy Metals",
        TRUE ~ Simplified_Etiology))
```

### Column for Location
```{r}
testing_data_fin

# Create Column
testing_data_fin <- testing_data_fin %>%
  rename(Location = Setting)%>%
  mutate(Location = case_when(
    grepl("Restaurant", Location) ~ "Restaurant", 
    TRUE ~ Location)) %>%
    separate_rows(Location, sep = "; ")

# Match Training Data Format
testing_data_fin <- testing_data_fin %>%
  mutate(Location = case_when(
    grepl("Private Home/Residence", Location, ignore.case = TRUE) ~ "Private Home/Residence",
    grepl("Caterer", Location, ignore.case = TRUE) ~ "Catering Service",
    grepl("Office/Indoor Workspace", Location, ignore.case = TRUE) ~ "Office/Indoor Workspace",
    grepl("Banquet", Location, ignore.case = TRUE) ~ "Banquet Facility",
    grepl("Fair", Location, ignore.case = TRUE) ~ "Fair/Festival",
    grepl("Religious", Location, ignore.case = TRUE) ~ "Religious Facility",
    grepl("Grocery Store", Location, ignore.case = TRUE) ~ "Grocery Store",
    grepl("School/College/University", Location, ignore.case = TRUE) ~ "School/College/University",
    grepl("Prison/Jail", Location, ignore.case = TRUE) ~ "Prison/Jail",
    grepl("Long-Term Care", Location, ignore.case = TRUE) ~ "Nursing Home/Assisted Living Facility",
    grepl("Child", Location, ignore.case = TRUE) ~ "Child Daycare",
    grepl("Farm/Dairy", Location, ignore.case = TRUE) ~ "Farm/Dairy",
    grepl("Hotel/Motel", Location, ignore.case = TRUE) ~ "Hotel/Motel",
    TRUE ~ Location))
```

## Ensure Training and Test Data Have Same Columns

To ensure compatibility for evaluating model or machine learning performance between the two datasets, column names must be the same.

```{r}
colnames(training_data_fin)
colnames(testing_data_fin)

training_data_fin <- training_data_fin %>%
  select(Year, State, Month, Numeric_Month, State, Location, 
         Food, `IFSAC Category`, 
         Illnesses, Hospitalizations, Pathogen_Type, Simplified_Etiology)

testing_data_fin <- testing_data_fin %>%
  rename(Food = `Food Vehicle`) %>%
  select(Year, State, Month, State, Location, 
         Food, `IFSAC Category`, 
         Illnesses, Hospitalizations, Pathogen_Type, Simplified_Etiology)

colnames(training_data_fin)
colnames(testing_data_fin)
```

# Drop Rows Containing NA
```{r}
training_data_final <- training_data_fin %>%
  mutate(Simplified_Etiology = na_if(Simplified_Etiology, "")) %>%
  drop_na()
testing_data_final <- testing_data_fin %>%
  mutate(Simplified_Etiology = na_if(Simplified_Etiology, "")) %>%
  drop_na()
```
